\documentclass[a4paper]{article}

\usepackage{a4wide}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}

\newcommand{\first}[1]{\mathsf{first}({#1})}
\renewcommand{\flat}[1]{\overline{#1}}
\newcommand{\last}[1]{\mathsf{last}({#1})}
\newcommand{\prearrow}[1]{{\rightarrow_{#1}^{-1}}}

\renewcommand{\iff}{\quad \mbox{iff} \quad}


\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{definition}[theorem]{Definition}

\title{The lifting of patterns}

\author{Johannes Marti and Leif Sabellek}

\begin{document}

%\maketitle


\section{The definition}

Let $P$ be a pattern. The lifting $L(P)$ is a pattern that has as nodes
all subsets $V \subseteq P$ with $V \neq \emptyset$ and $|V| \leq 2$.
The relations are defined such that for both $a \in \{0,1\}$ we have
\[
 V \rightarrow_a W \iff \mbox{there is some } v \in V \mbox{ such that }
v \rightarrow_a w \mbox{ for all } w \in W.
\]
One can show that if $h : P \to Q$ is a homomorphism then $L(h) : L(P)
\to L(Q)$ with $L(h)(\{u,v\}) = \{h(u),h(v)\}$ is a homomorphism. Thus,
the lifting is a monotone operation on the homomorphism lattice. One can
also show that there is a homomorphism $e : P \to L(P)$ for all patterns
$P$. The homomorphism $e$ maps any $v \in P$ to $e(v) = \{v\}$.

The motivation for this notion is that one can show the following: For
all patterns $P$ and numbers $n$:
\begin{equation} \label{eq:lifting property}
 T_{n + 1} \to P \iff T_n \to L(P).
\end{equation}
Thus, it follows that $T_n \to P$ iff $T_0 \to L^n(P)$, where $L^n$
denotes the $n$-fold application of $L$. Note that $T_0 \to L^n(P)$ just
means that $L^n(P)$ has a point with reflexive $0$ and $1$-loops.


\section{The main property}

We are now going to prove \eqref{eq:lifting property}.
\begin{theorem}
 For all patterns $P$ and numbers $n \in \omega$
\begin{equation*}
 T_{n + 1} \to P \iff T_n \to L(P).
\end{equation*}
\end{theorem}
\begin{proof}
 For the left-to-right direction assume that we have a homomorphism $h :
T_{n + 1} \to P$. We define that homomorphism $g : T_n \to L(P)$ by
setting $g(v) = \{h(v0),h(v1)\}$ for all $v \in \Sigma^n$. To see that
this defines a homomorphism assume that we have $v \rightarrow_b w$ in
$T_n$ for $v,w \in \Sigma^n$. We need to show that $\{h(v0),h(v1)\}
\rightarrow_b \{h(w0),h(w1)\}$ holds in $L(P)$. Consider $bw$ in $T_{n +
1}$. By definition of the de Bruijn graph $T_{n + 1}$ we have that $bw
\rightarrow_b w0$ and $bw \rightarrow_b w1$. It thus suffices to show
that there is some $c \in \Sigma$ for which $bw = vc$. If $n = 0$ then
we can simply set $c = b$. If $n > 0$ then it follows from $v
\rightarrow_b w$ that $v = b v'$ and $w = v' d$ for some $v' \in
\Sigma^{n - 1}$ and $d \in \Sigma$. We then have that $b v' d
\rightarrow_b w 0$ holds in $T_{n + 1}$. Because the $b$-predecessor of
$w 0$ is unique it follows that $b w = b v' d = v d$. So we can set $c =
d$.

 For the right-to-left direction assume that there is a homomorphism $g
: T_n \to L(P)$. If $n = 0$ then $h : T_0 \rightarrow L(P)$ yields a
point $h(\star) = \{x,x'\}$ in $L(P)$ such that $\{x,x'\} \rightarrow_b
\{x,x'\}$ for both $b \in \Sigma$. This means that for both $b \in
\Sigma$ there is an $x_b \in \{x,x'\}$ such that $x_b \rightarrow_b x_c$
for both $c \in \Sigma$. The homomorphism $h : T_1 \to P$ can then map
$0$ to $x_0$ and $1$ to $x_1$. In the other case we have that $n > 0$.
To define a homomorphism $h : T_{n + 1} \to P$ consider any $w c$ in
$T_{n + 1}$, where $w \in \Sigma^n$ and $c \in \Sigma$. Because $n > 0$
we have $w = b u$ for some $b \in \Sigma$ and $u \in \Sigma^{n - 1}$. In
$T_n$ we then get the edge $b u \rightarrow_b u c$ and because $g$ is a
homomorphism there must be a $b$-edge from $g(w) = g(bu) = \{x,x'\}$ to
$g(uc) = \{y,y'\}$. Let $x^\circ \in \{x,x'\}$ be such that $x^\circ
\rightarrow_b y$ and $x^\circ \rightarrow_b y'$ both hold in $P$. We
define $h(wc) = x^\circ$. To see that $h : T_{n + 1} \to P$ is a
homomorphism consider an edge $b u c \rightarrow_b u c d$ in $T_{n + 1}$
for any $u \in \Sigma^{n - 1}$ and $b,c,d \in \Sigma$. By the definition
of $h$ we know that $h(u c d)$ must be one of the elements from $g(u c)
= \{y,y'\}$. But $h(b u c)$ was defined to be the $x^\circ$ such that
$x^\circ \rightarrow_b y$ and $x^\circ \rightarrow_b y'$. Hence, $h(b u
c) \rightarrow_b h(u c d)$.
\end{proof}


\section{More facts}

One can check that $L$ defines a functor.

\begin{definition}
 Let $P$ be any pattern. Define the \emph{canonical embedding} $e : P
\to L(P)$ such that $e(x) = [x]$. Inductively we can define $e^n : P \to
L^n(P)$ such that $e^0 : P \to P$ is the identity on $P$ and $e^{n + 1}
= e \circ e^n$.
\end{definition}

One can check that $e^{n + m} = L^n(e^m) \circ e^n$ for all $n$ and $m$.


\begin{definition}
 Define the \emph{deepening} $d : L^1(P) \to L^2(P)$ such that
$d(\{x,y\}) = \{\{x\},\{y\}\}$.
Define the \emph{deepening} $d^m_n : L^n(P) \to L^{m +
n}(P)$
\end{definition}

\begin{proposition}
Let $P$ be any pattern. Then the following hold:
\begin{enumerate}
 \item $x \rightarrow_a y$ iff $e(x) \rightarrow_a e(y)$ for all $x,y
\in P$,
 \item $x \rightarrow_a y$ iff $d^m_n(x) \rightarrow_a d^m_n(y)$ for all
$x,y \in L^n(P)$,
% \item $x \rightarrow_a y$ implies $x \rightarrow_a e(y)$,
\end{enumerate}
\end{proposition}


\section{More thoughts}

We can also restrict our attention to the subcategory of all patterns in
which every point has at least one $a$-predecessor for every label $a$.
$L$ can then be constraint to only add doubletons if they have this
property. In the following we assume to be in this subcategroy, and
write $E$ for the $L$ of the outer category, which does not have this
restriction.

\begin{lemma}
 Let $X \in E^k(P)$ and $U,V \in L^n(P)$ such that
\[
 \flat{X} \subseteq [\flat{U}]{\rightarrow_0} \cap
[\flat{V}]{\rightarrow_1}.
\]
Then there is some $i$ such that $U \rightarrow_0 d^i(X)$ and $V
\rightarrow_1 d^i(X)$.
\end{lemma}

\begin{lemma}
 If $P$ is construction deterministic then there is a $k$ and a $U \in
L^k(P)$ such that $\flat{U} = P$. Moreover, for all $X \in E^i(P)$ there
is some $j$ such that $d^j(X) \in L^{i + j}(P)$.
\end{lemma}
\begin{proof}
 It follows that $[\flat{U}]{\rightarrow_0} = P =
[\flat{U}]{\rightarrow_1}$ and thus 
\end{proof}


\end{document}
